Soil example
========================================================

```{r}
library(RandomFields)
library(reshape2)
library(ggplot2)
theme_set(theme_bw())
data(soil)
pts <- soil[,c(1,2)]
d <- soil$moisture
```

Define some graphical parameters first
```{r}
ggplot(soil,aes(x=x.coord,y=y.coord,colour=moisture))+
  geom_point(size=4,alpha=0.7)+
  scale_colour_gradient(low="blue",high="red")
```

Quick look at $x$-$y$ trends:
```{r}
m1 <- melt(subset(soil,select=c(x.coord,y.coord,moisture)),
     id.var="moisture")
ggplot(m1,aes(x=value,y=moisture))+
  geom_point()+geom_smooth(method="loess")
```

Exploring the trend in $y$ a bit more:
```{r}
ggplot(soil,aes(x=y.coord,y=moisture))+
  geom_point()+
  geom_smooth(method="loess")+
  geom_smooth(method="gam",colour="red")+
  geom_smooth(method="lm",formula=y~poly(x,2),
              colour="orange")+
  stat_summary(fun.y=mean,geom="line",colour="purple")
```

## empirical variogram
```{r}
maxbin <- max(dist(pts)) / 2
ev <- EmpiricalVariogram(pts, data=d, grid=FALSE,
                         bin=c(-1,seq(0,maxbin,l=30)))
(g0 <- ggplot(data.frame(ev),aes(x=centers,y=emp.vario))+
  geom_point(aes(size=n.bin),alpha=0.5))
```

```{r}
g0 + geom_ribbon(aes(ymin=emp.vario-2*sd,ymax=emp.vario+2*sd),alpha=0.2)
```

Fit parameters of the Whittle-Matérn model (a family that includes exponential and Gaussian models) by MLE:

```{r}
fit <- fitvario(x=pts, data=d, model="whittle",
  par=rep(NA,5),
  mle.m="ml", cross.m=NULL)
str(fit)
```
## plot the fitted model and the empirical variogram

```{r}
plot(ev$c, ev$emp.var, ylim=c(0,11), ylab="variogram", xlab="lag")
gx <- seq(0.001, max(ev$c), l=100)
if(!is.null(by.eye)) lines(gx, Variogram(gx, model=by.eye)) 
lines(gx, Variogram(gx, model=fit$ml$model), col=2)
lines(gx, Variogram(gx, model=fit$plain$model), col=3)
lines(gx, Variogram(gx, model=fit$sqrt.nr$model), col=4)
lines(gx, Variogram(gx, model=fit$sd.inv$model), col=6, lty=2)
legend(120, 4, c("empirical", "by eye", "ML", "lsq", "sqrt(n) lsq",
               "sd^-1 lsq"),
       lty=c(-1, rep(1, 5)), pch=c(1, rep(-1, 5)),
       col=c(1, 1, 2, 3, 4, 6), cex=1.4)
```

## map of expected values

```{r}
k <- Kriging("O", x=x, y=y, grid=TRUE, model=fit$ml$model, given=pts, data=d)
par(mfrow=c(1,2))
plot(pts, col=colour[1+(cn-1)*((d-zlim[1])/diff(zlim))],
     pch=16, xlab="x [cm]", ylab="y [cm]")
my.legend(lu.x, lu.y, zlim=zlim, col=colour, cex=1)
image(x, y, k, col=colour, zlim=zlim, xlab="x [cm]", ylab="y [cm]")
par(bg="white")
```


## what is the probability that at no point of the
## grid given by x and y the moisture is greater than 24 percent?

```{r}
RFparameters(Print=1, CE.force=FALSE, CE.trials=3, CE.useprimes=TRUE)
cs <- CondSimu("O", x=x, y=y, grid=TRUE, model=fit$ml$model, given=pts,
               data=d, n=10) # better n=100 or n=1000
par(mfcol=c(2,3))
plot(pts, col=colour[1+(cn-1)*((d-zlim[1])/diff(zlim))], pch=16,
     xlab="x [cm]", ylab="y [cm]", cex.axis=1.5, cex.lab=1.5)
my.legend(lu.x, lu.y, zlim=zlim, col=colour, cex=0.5)
image(x, y, k, col=colour, zlim=zlim, xlab="x [cm]", ylab="y [cm]")
for (i in 1:4)
  image(x, y, cs[, , i], col=colour, zlim=zlim,
        xlab="x [cm]", ylab="y [cm]") 

mean(apply(cs<=24, 3, all)) ## about 40 percent ...
```
