Spatial analysis
=========================================

```{r opts,echo=FALSE}
opts_chunk$set(tidy=FALSE,fig.width=4,fig.height=6)
```

Huge topic!

## Key references

Diggle (point patterns); Cressie (everything); Diggle and Ribeiro (geostatistics); Dormann et al (GLMMs for species presence/abundance); Haining; [@pinheiro_mixed-effects_2000]

### Point processes
 
* just looking at the distribution of ("unmarked") points, asking whether they are clustered, random, regular ("overdispersed" [!!])
* standard summary: Ripley's $K$ (number of points within radius $r$ of a randomly chosen point, divided by the overall density); $L=\sqrt{K/\pi}$ should be linear ...
    * have to deal with *edge corrections*: corrected estimators, null distributions via permutation tests   
* tests of *complete spatial randomness* (CSR)

### Lattices
I don't have much to say here: data are often *sampled* this way but we more typically model them in continuous space, or on a graph

### Graphs/networks

* Really more general than space: don't even need to satisfy "spatial" properties (e.g. could be a social network rather than a spatial graph)
* different ways to represent spatial networks
    * neighbor list (with weights)
    * adjacency matrix (weighted)
* Deriving weights matrix $W$ from spatial data  (from [Bannerjee presentation](http://www.unc.edu/~rls/s940/Areal.pdf):
    * =1 if nearest neighbor (or $n^{\text{th}}$ nearest neighbor?), 0 otherwise
    * polygons: "neighbor"="share a boundary", then as above?
    * =1 if distance < threshold
    * inverse-distance weighted (cutoff beyond some distance to make the matrix *sparse*?)
    * exponential weighting (but need to choose decay parameter ...)
    * $W$ doesn't need to be symmetric
* Voronoi diagrams/Delaunay/Dirichlet tesselations

### Random fields
* Random fields
   * Point samples of a continuously varying field
   * *Gaussian* random fields (multivariate normal with specified spatial correlation function)
   * build non-Gaussian random fields on top of Gaussian RF; hierarchical models

## Trend vs correlation

* stationarity, isotropy
* large- vs small-scale patterns
* mean models vs variance models
* (fitting small-scale spatial pattern via splines)
   
## Not-really-spatial models

Two kinds of models that I don't classify as spatial models:

* Models where the samples are taken spatially (i.e. measuring diversity vs rainfall from a bunch of plots, or environment and community samples in many plots (ordination etc.), but we just use space as a grouping factor, not considering which plots are closer to each other
* As above, but with $x$/$y$ (lat/long, eastings/northings etc.) included as input
variables, possibly with quadratic terms (`poly(x,y,degree=2)`) - in spatial statistics this is called
*trend surface analysis*.
* in other words, truly spatial analyses take spatial *relationships* among points into account

## Avoiding spatial analysis

* Non-spatial analysis; show that residual pattern is insignificant, biologically and statistically (maps, or e.g. Moran's $I$)
* Aggregate data (buffering etc.) until aggregated observations are approximately independent
* Claim that spatial correlations don't bias your estimates (true for *linear* models) and that the adjustment to the confidence intervals is not important (McGill)
* Dutilleul's method?

## Spatial diagnostics

* graphical: maps of residuals (e.g. size=absolute magnitude, red vs blue = positive/negative, or diverging color scale)
* semi-graphical: *semivariogram* or *autocorrelation function*

## Analyses based on weight matrices

* Parallel with Ives and Zhu's "generalized least squares" example: correlation matrix assumed known
* Moran's $I$ (analogue of lagged autocorrelation), Geary's $C$

* Assume we are willing to specify the weight matrix $W$ *a priori*
* Efficient matrix-based solutions: [Conditional and simultaneous autoregression](http://stats.stackexchange.com/questions/277/spatial-statistics-models-car-vs-sar):
    * *Non-spatial model*: my house value is a function of my home gardening investment.
    * *Conditional autoregression*: my house value is a function of the gardening investment of my neighbours.
    * *Simultaneous autoregression*: my house value is a function of the house values of my neighbours.

## Geostatistical models

### Correlation models

(Semi)variance: $S(r_{ij})=(x_i-\bar x)(x_j-\bar x)/2$.

* starts at the *nugget*; continues out to the *sill*
* Useful for exploration (mostly not for model fitting nowadays)
* Usually makes a giant, uninterpretable point cloud
unless one bins the data or fits some kind of smooth curve

```{r echo=FALSE,fig.width=8,message=FALSE}
par(las=1,bty="l")
par(mfrow=c(1,2))
library(RandomFields)
param1 <- c(mean=0,variance=2,nugget=0.5,scale=1,nu=1/2)
param2 <- c(mean=0,variance=2,nugget=0.5,scale=1,nu=2)
param3 <- c(mean=0,variance=2,nugget=0.5,scale=1,nu=10000)
curve(Variogram(x,model="whittle",param=param1),from=0,
      to=5,ylab="variogram",n=101)
curve(Variogram(x,model="matern",param=param2),
      col=2,add=TRUE)
curve(Variogram(x,model="matern",param=param3),
      col=4,add=TRUE)
legend("bottomright",
       lty=1,col=c(1,2,4),
       c("exponential (s=0.5)",
         "s=2",
         "s=10000"))
curve(CovarianceFct(x,model="matern",param=param1),
      from=0,to=5,ylab="covariance",n=101)
curve(CovarianceFct(x,model="matern",param=param2),
      col=2,add=TRUE)
curve(CovarianceFct(x,model="matern",param=param3),
      col=4,add=TRUE)
```

```{r echo=FALSE,fig.width=10,fig.height=4}
xvec <- seq(0,5,length=101)
g1 <- GaussRF(x=xvec,y=xvec,grid=TRUE,model="matern",param=param1)
g2 <- GaussRF(x=xvec,y=xvec,grid=TRUE,model="matern",param=param2)
g3 <- GaussRF(x=xvec,y=xvec,grid=TRUE,model="matern",param=param3)
par(mfrow=c(1,3),mar=c(1,0.5,3,0.5))
image(g1,axes=FALSE,useRaster=TRUE,main="exponential")
image(g2,axes=FALSE,useRaster=TRUE,main="s=2")
image(g3,axes=FALSE,useRaster=TRUE,main="Gaussian")
```
* must obey constraints: *positive definiteness* (equivalent to 'no negative variances' or 'no impossible correlation geometries')
* typically use a small set of well-studied possibilities
    * classical: spherical, linear, exponential, Gaussian: each have a
    * newer: Matérn (includes exponential and Gaussian as special cases), powered exponential
    * all start at 1 (unless there's a *nugget effect*), decrease eventually to zero; most are positive everywhere
    * spatial *variogram* or *semivariogram*; equivalent information but easier to compute

* spatial prediction: *kriging*

## R packages

* `spdep`: weight matrices, Moran's $I$, CAR/SAR
* `RandomFields`: simulating Gaussian RF of all types
* `nlme`: `g[n]ls` and `[n]lme` can handle standard spatial autocorrelation structures (only within blocks)
* `ramps`: Bayesian MCMC fitting of geostatistical models.  Also lots of additional spatial correlation structures, including basing correlation on great-circle distances                      
* `geoR`: spatial LMs and GLMMs (but without additional grouping structures)
* `ape`: correlation classes for phylogenetic correlations
* `INLA`: complex but powerful package for spatial (among others) fitting

AD Model Builder [spatial ex.], BUGS (GeoBUGS)
