---
title: "Tidying data with tidyr"
author: "Karthik Ram"
date: "July 26, 2014"
output: html_document
---

`tidyr` is a new package in R that is meant to be an update/replacement to the widely used `reshape2` package. It works similar to `dplyr` in that it uses pipes (`%>%`) to chain the output from one function into the next. In addition to giving you the speed of dplyr, it also makes complex code human readable. Here are a few examples on how to use `tidyr`. I pulled together various examples from stackoverflow and answered them below.

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = " ",
  error = FALSE,
  message = FALSE,
  tidy = FALSE
)
```

```{r load libraries}
# http://stackoverflow.com/questions/7449198/
library(tidyr)
library(dplyr)
```

```{r example1}
set.seed(1001)
d <- expand.grid(f1=LETTERS[1:3],f2=letters[1:3],
                 f3=factor(as.character(as.roman(1:3))),rep=1:4)
d$y <- runif(nrow(d))
d$z <- rnorm(nrow(d))

results <- d %>% group_by(f1, f2, f3) %>% summarise(y.mean = mean(y), z.mean = mean(z))
head(results)
```

## Example 2

```{r ex2}
# http://stackoverflow.com/questions/23953163/
library(tidyr)
library(dplyr)

mydata <- data.frame(var = rep(c("A","B","C"),each = 3), 
	code = rep(c("x","y","z"),3), 
	yearA = 1:9, yearB = 10:18, yearC = 20:28)

mydata %>% gather(year, value, yearA:yearC) %>%
mutate(var = paste0("var", ".", var)) %>%
spread(var, value)	

```

## Example 3

```{r ex3}
# http://stackoverflow.com/questions/23945350/
library(tidyr)
library(dplyr)

dw <- read.table(header=T, text='
 sbj f1.avg f1.sd f2.avg f2.sd  blabla
   A   10    6     50     10      bA
   B   12    5     70     11      bB
   C   20    7     20     8       bC
   D   22    8     22     9       bD
 ')

dw %>% gather(v, value, f1.avg:f2.sd) %>% 
separate(v, c("var", "col")) %>% 
arrange(sbj) %>% 
spread(col, value)

dw %>% gather(v, value, f1.avg:f2.sd) %>% 
separate(v, c("var", "col")) %>% 
spread(col, value)

```

