## `rgbif` use case No. 2 - Biodiversity comparisons across cities

In this example, we compare biodiversity in different regions using GBIF data from rgbif. This example can be done using BISON data as well with our rbison package.

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6, 
  fig.height = 5
)
```

### Load libraries

```{r load}
library(rgbif)
library(ggplot2)
library(plyr)
```


### Get GBIF data using the rgbif library.

We wont not restrain our search to any particular taxonomic group, although you will likely do that in your own research. We will first make a new column with single lat/long coordinates for each cell for easy plotting. Finally we select 100 random cell IDs.

```{r getbounding}
temp <- gbifdata(densitylist(originisocountrycode = "US"))
temp <- transform(temp, lat = (minLatitude+maxLatitude)/2, lon = (minLongitude + maxLongitude) / 2) 
cellids <- sample(temp$cellid, 100)
```

Then search for data for each of those cell IDs. We then define a function to pass in each cell ID.

```{r getdata, cache = TRUE}
getdata <- function(x, maxresults = 100) {
  results <- occurrencelist(cellid = x, coordinatestatus = TRUE, maxresults = maxresults)
  df <- gbifdata(results, coordinatestatus = TRUE)
  data.frame(cellid = x, richness = length(unique(df$taxonName)))
}

results_with_locations <- ldply(cellids, getdata)
merged_results <- merge(results_with_locations, temp[, c('cellid', 'lat', 'lon')], by = 'cellid')
# remove points outside the US
final_results <- merged_results[merged_results$lat < 49 & merged_results$lat > 24.7433195 & merged_results$lon > -130 & merged_results$lon < -66.9513812, ]
```

### Plot data

```{r plot}
mapp <- map_data('state')
ggplot(mapp, aes(long, lat)) +
  geom_polygon(aes(group = group), fill = "white", alpha = 0, color = "gray80", size = 0.8) +
  geom_point(data = final_results, aes(lon, lat, color = richness), size = 3, alpha = 0.8) +
  scale_color_gradient2(low = "white", mid="lightblue", high = "blue") +
  labs(x = "", y = "") +
  theme_bw(base_size = 14) + 
  theme(legend.position = "bottom", legend.key = element_blank())
```

### Notes

Bounding lat/long data from [here](https://raw.github.com/amyxzhang/boundingbox-cities/master/boundbox.txt)