## rbison use case - Macroecology: testing the species-abundance distribution

Keep in mind that BISON data is at least in part within GBIF, which can be accessed from R using our rgbif package. However, BISON provides a slightly different interface to their data than does GBIF, and if you are just interested in US data, then BISON may be easier to use. 

In addition, this example can be done using GBIF data.

In this example, we do some preliminary work in exploring species-abundance distribution. 

```{r defaults, eval=TRUE, echo=FALSE, cache=TRUE}
opts_chunk$set(comment=NA, warning=FALSE, message=FALSE, fig.width=7, fig.height=6)
```

### Load libraries

```{r load}
library(rbison)
library(ggplot2)
library(plyr)
library(doMC)
```

### Get BISON data using the rOpenSci package rbison.

We'll not restrain our search to any particular taxonomic group, although you will likely do that in your own research. We'll make a new column with single lat/long coordinates for each cell for easy plotting. Last, we'll select 100 random cell IDs.

```{r getdata}
mynames <- c("Helianthus annuus", "Pinus contorta", "Poa annua", "Madia sativa", "Arctostaphylos glauca", "", "Heteromeles arbutifolia", "Symphoricarpos albus", "Ribes viburnifolium", "Diplacus aurantiacus", "Salvia leucophylla", "Encelia californica", "Ribes indecorum", "Ribes malvaceum", "Cercocarpus betuloides", "Penstemon spectabilis")

getdata <- function(x, count=2){
  out <- bison(species=x, county = "Los Angeles", count=count)
  bison_data(out, "data_df")
}

registerDoMC(cores = 4)
out <- ldply(mynames, getdata, count=500, .parallel=TRUE)
df <- ddply(out, .(name), summarise, abd = length(name))
```

### Rank-abundance plot, with fits of various models 

```{r plot}
library(vegan)
plot(radfit(df$abd))
```

### Further reading

Read more about plotting abundance distributions [here](http://cran.r-project.org/web/packages/vegan/vignettes/diversity-vegan.pdf). 