Classification and Regression Tree Example
----------------------------------------------

```{r opts,echo=FALSE}
## internal bookkeeping: you may ignore this
opts_chunk$set(results="hide",tidy=FALSE,fig.height=4,
              fig.width=6)
knit_hooks$set(treefig=function(before, options, envir) {
     par(xpd=NA,cex=0.9)})
```

## Packages used
```{r pkglist,echo=FALSE}
usedpkgs <- sort(c("rpart","ade4","mvpart","vegan"))
i1 <- installed.packages()
print(i1[usedpkgs,"Version"],quote=FALSE)
```

## Data

```{r rpart,message=FALSE}
library(rpart)
summary(cu.summary)
```

(note: be careful *not* to load the `mvpart` package in this section -- it will load a different version of the `rpart()` function)

## Grow tree

```{r}
(fit <- rpart(Mileage~Price + Country + Reliability + Type,
        method="anova", xval =100, data=cu.summary))
```
(if you know you want to use all the variables other than
the response variable as input variables,
`rpart(response~.,...)` will work)
Note: If your data are binary and you are interested in performing a classification tree, method = “class”

```{r treefig=TRUE}
printcp(fit) # display the results
plotcp(fit) # visualize cross-validation results
summary(fit) # detailed summary of splits
```

# plot
```{r treefig=TRUE}
plot(fit)
text(fit)
```

## prune

```{r}
cpval <- fit$cptable[which.min(fit$cptable[,"xerror"]),"CP"]
pfit<- prune(fit, cp=cpval)
```

```{r  treefig=TRUE}
plot(pfit)
text(pfit)
summary(pfit)
```

# Pick your own tree size

```{r}
dfit <- rpart(Mileage~., method="anova", 
              maxdepth=2, data=cu.summary)
```

```{r  treefig=TRUE}
plot(dfit)
text(dfit)
summary(dfit)
```

(Code modified from [Quick-R: Accessing the Power of R](http://www.statmethods.net))

# Multivariate Regression Tree Example

## Data 

[Data Information](http://www.inside-r.org/packages/cran/ade4/docs/doubs)

```{r ade4,message=FALSE}
library(ade4)
data(doubs)
env=doubs$env
spe = doubs$fish
```

## Transform response variables

The transformation consists of expressing each fish density as a proportion of the sum of all densities in the analytical unit and taking the square root of the resulting value (Legendre and Gallagher 2001).The square-root portion of the transformation decreases the importance of the most abundant species.

```{r vegan,message=FALSE}
library(vegan)
spe.norm=decostand(spe, "hellinger")
```

## Multivariate Regression Tree 

```{r mvpart,warning=FALSE,message=FALSE}
library(mvpart)
spe.ch.mvpart<-mvpart(data.matrix(spe.norm) ~., env,  
                      xv="1se", xval=nrow(spe), 
                      xvmult=100, which=4)
summary(spe.ch.mvpart)
printcp(spe.ch.mvpart)
```

Or we can use `pick`
```{r eval=FALSE}
spe.ch.mvpart<-mvpart(data.matrix(spe.norm) ~., env,  
                      xv="pick", xval=nrow(spe), 
                      xvmult=100, which=4)
summary(spe.ch.mvpart)
printcp(spe.ch.mvpart)
```

* `xv` = Selection of tree by cross-validation: 
    * `"1se"` - gives best tree within one SE of the overall best,
    * `"min"` - the best tree
    * `"pick"` - pick the tree size interactively, 
    * `"none"` - no cross-validation.
* `xval` = Number of cross-validations or vector defining cross-validation groups (here we use as many rows there are in the dataset because it is a small dataset)
* `xvmult` = Number of multiple cross-validations.
* `which` = Which split labels and where to plot them, 1=centered, 2 = left, 3 = right and 4 = both.

(Modified R Code from “Numerical Ecology with R” by Borcard et al. 2012

